---
title: "The Paired t-test"
output: pdf_document
---

# Overview
This script simply provides some additional materials that were used to make the power point presentation, such as TeX equations and illustrative figures.

# Script preamble
```{r preamble, eval=TRUE,echo=TRUE,results='hide'}
# Preamble - load packages and save colors for plotting
library(tidyverse)


# bright color blind palette:  https://personal.sron.nl/~pault/ 
cblack <- "#000004FF"
cblue <- "#3366AA"
cteal <- "#11AA99"
cgreen <- "#66AA55"
cchartr <- "#CCCC55"
cmagent <- "#992288"
cred <- "#EE3333"
corange <- "#EEA722"
cyellow <- "#FFEE33"
cgrey <- "#777777"

```



# Notation used in the presentation

The mean within-pair difference, $\mu_D$, is estimated as:

\begin{displaymath}
\hat{\mu}_D = \frac{1}{n} \sum_i^n D_i
\end{displaymath}

An estimate of the standard deviation of $D$ is:

\begin{displaymath}
\hat{S}_D^2 = \frac{1}{n-1} \sum_i^n (D_i - \hat{\mu}_D)^2 
\end{displaymath}

With standard error:

\begin{displaymath}
SE(\hat{\mu}_D) = \frac{\hat{S}_D}{\sqrt{n}}
\end{displaymath}


A t statistic based on the paired differences is:

\begin{displaymath}
t = \frac{\hat{\mu}_D}{(\hat{S}_D / \sqrt{n}) } =  \frac{\hat{\mu}_D}{SE(\hat{\mu}_D)}
\end{displaymath}


Comparing the paired t test with a two sample t test.

If the variances of $Y_C$ and $Y_I$ are equal ($\sigma$), the variance of their paired difference, $D$ is:

\begin{displaymath}
\sigma_{D}^2 = 2 \sigma^2 (1-r)
\end{displaymath}

In an unpaired study, where $Y_C$ and $Y_I$ are independent samples, again with equal variance, the variance of their difference is the sum of their variances:


\begin{displaymath}
\sigma_{Y_I - Y_C}^2 = \sigma^2 + \sigma^2 = 2 \sigma^2
\end{displaymath}


# WHO standard normal distribution

```{r who zscores}
qz <- seq(-5,5,by=0.01)
dz <- dnorm(qz,mean=0,sd=1)
dstd <- data.frame(qz,dz)

pz <- ggplot(data=dstd,aes(x=qz,y=dz)) +
  geom_line() +
  scale_x_continuous(breaks=-3:3) +
  coord_cartesian(xlim=c(-3,3)) +
  labs(x="WHO Length-for-age Z score (LAZ)",y="Density") +
  theme_minimal(base_size=16)
pz
ggsave(filename="~/washb-pairedttest/results/who-laz.png",plot=pz,device="png")


pzs <- pz +
  scale_x_continuous(breaks=-5:4) +
  coord_cartesian(xlim=c(-5,4)) +
  geom_vline(xintercept=-2,lty="dashed") +
  geom_vline(xintercept=-3,lty="dashed") +
  annotate("text",x=-2.5,y=0.41,label="Stunted") +
  annotate("text",x=-4,y=0.41,label="Severely stunted")

pzs
ggsave(filename="~/washb-pairedttest/results/who-laz-labeled.png",plot=pzs,device="png")
```


```{r load bangladesh data}
# load and merge the final analysis files
# treatment assignments, enrollment charactersitics, and anthropometry measurements
d_tr    <- read.csv("~/washb-pairedttest/data/washb-bangladesh-tr-public.csv")
d_anth  <- read.csv('~/washb-pairedttest/data/washb-bangladesh-anthro-public.csv')


#  merge the analysis files, loaded above
d <- left_join(d_tr,d_anth,by=c("clusterid","block"))

# filter to children with measurements at year 2 (final endpoint)
# with valid LAZ measuresments (laz_x==0)
# filter to the control arm
dc <- d %>%
  filter(svy==2 & laz_x==0 & tr %in% c("Control")) %>%
  select(block,tr,dataid,length,laz) %>%
  mutate(tr=droplevels(tr))

```

```{r mean and sd estimates}

summary(select(dc,length,laz))

dsum <- dc %>% select(length,laz) %>% summarize(sd_len=sd(length),sd_laz=sd(laz))
dsum


```

```{r figures of length and laz}

plength <- ggplot(data=dc,aes(x=length)) +
  geom_density(fill=cblue,color=NA,alpha=0.5) +
  labs(x="Length in centimeters") +
  annotate("text",x=73,y=0.11,label="Bangladesh Trial\nControl Group",color=cblue) +

  theme_minimal(base_size = 16)

plength
ggsave(filename="~/washb-pairedttest/results/washb-control-cm.png",plot=plength,device="png")



plaz <- ggplot(data=dc,aes(x=laz)) +
  geom_density(fill=cblue,color=NA,alpha=0.5) +
  geom_line(data=dstd,aes(x=qz,y=dz))+
  scale_x_continuous(breaks=-5:4) +
  coord_cartesian(xlim=c(-5,3))+
  labs(x="Length-for-age Z-score (LAZ)") +
  annotate("text",x=2,y=0.38,label="WHO Reference Standard") +
  annotate("text",x=-3.5,y=0.38,label="Bangladesh Trial\nControl Group",color=cblue) +
  theme_minimal(base_size = 16)

plaz
ggsave(filename="~/washb-pairedttest/results/washb-control-laz.png",plot=plaz,device="png")

```


# Null distribution
Sampling distribution of the mean, under the null
```{r plot the t distribution}
# normal distribution with mean 0 and SD = 0.053 (the SE of the differences observed in the Bangladesh trial)
qt <- seq(-5,5,by=0.01)
dent <- dt(qt,df=89)
datt <- data.frame(qt,dent)

pt89 <- ggplot(data=datt,aes(x=qt,y=dent)) +
  geom_line() +
  scale_x_continuous(breaks=-5:5) +
  coord_cartesian(xlim=c(-5,5)) +
  labs(x="t distribution (df=89)",y="Density") +
  theme_minimal(base_size=16)
pt89

ggsave(filename="~/washb-pairedttest/results/tdist-89df.png",plot=pt89,device="png")


```

# t distribution

```{r plot the null distribution}
# t distribution with 89 degrees of freedom
qnull <- seq(-0.3,0.3,by=0.001)
nullden <- dnorm(qnull,mean=0,sd=0.05312652)
dnull <- data.frame(qnull,nullden)

pnull <- ggplot(data=dnull,aes(x=qnull,y=nullden)) +
  geom_line() +
  scale_x_continuous(breaks=round(seq(-0.3,0.3,by=0.1),2)) +
  coord_cartesian(xlim=c(-0.3,0.3)) +
  labs(x="Distribution of pair-wise differences under the null",y="Density") +
  theme_minimal(base_size=16)
pnull

ggsave(filename="~/washb-pairedttest/results/washb-nulldist.png",plot=pt89,device="png")


```


# Student's first t-test

```{r student data}
m1 <- c(0.7,-1.6,-0.2,-1.2,-0.1,3.4,3.7,0.8,0,2)
m2 <- c(1.9,0.8,1.1,0.1,-0.1,4.4,5.5,1.6,4.6,3.4)
dstudent <- data.frame(id=1:10,m1,m2) %>%
  mutate(diff=m2-m1)

summary(dstudent)
apply(dstudent,2,sd)
print(dstudent)


```
```{r plot student data}
dlong <- dstudent %>%
  select(id,m1,m2) %>%
  gather(key="measure",value="m",-id) %>%
  mutate(measure=factor(ifelse(measure=="m1","Dextro","Laevo")))

pstrip <- ggplot(data=dlong,aes(x=measure,y=m,group=id)) +
  geom_point() +
  labs(y="Additional hours slept",x="") +
  theme_minimal(base_size=18)

pstrip
ggsave(filename="~/washb-pairedttest/results/student-strip.png",plot=pstrip,device="png")

cbcols <- c(cblack,cred,corange,cyellow,cchartr,cgreen,cteal,cblue,cmagent,cgrey)

pline <- pstrip +  
  aes(color=factor(id)) + 
  geom_line() +
  scale_color_manual(values=cbcols,guide=guide_legend(title="Patient")) +
  theme_minimal(base_size=18) +
  theme(legend.position="none")

pline
ggsave(filename="~/washb-pairedttest/results/student-line.png",plot=pline,device="png")


```
```{r student ttests}

t.test(dstudent$m1,dstudent$m2,paired=FALSE)

t.test(dstudent$m1,dstudent$m2,paired=TRUE)

student_cor <- cor.test(dstudent$m1,dstudent$m2)
print(student_cor)

# relative efficiency:
print(1/(1-student_cor$estimate))

```

# Session Info
```{r session info}
sessionInfo()
```




